---
title: "PS270 Final Project"
author: "Arwen Lyons "
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(error = TRUE)
```



 -------------------- ----------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
`County`              Name of Pivot County 

`state`               State where county resides

`status.2020`         labeled retained when counties kept their voting record,
                      ie: Trump in 2016 and Trump in 2020 or Trump in 2016 and 
                      Biden in 2020, counties that switched votes are labeled
                      boomerang

`MOV.2020`            The winning parties margin of victory in 2020

`Trump.MOV.2016`      Republican Candidate Donald Trump's margin of victory in
                      2016

`Obama.MOV.2012`      Democratic presidential incumbent Barack Obama's margin 
                      of victory in 2012

`Obama_MOV.2008`      Democratic Candidate Obama's margin of victory in 2008

`Total`               Total National unemployment rate as reported by the 
                      Beauro of Labor Statistics in a given election year
                      
 ------------------------------------------------------------------------------
 
## UNEMPLOYMENT RATES
```{r}
library(tidyverse)
library(broom)
library(lubridate)
library(ggplot2)
library(dplyr)

nat_unemployment <- read_csv("National Unemployment Statistics sep 2004 - sept 2024 - Sheet1.csv")

nat_unemployment <- nat_unemployment|>
   select(Month, Total)

nat_unemployment <- nat_unemployment |>
  mutate(Month = as.numeric(format(as.Date(paste0("01", Month), format = "%d %b %Y"), "%Y%m")))
nat_unemployment


annual_avg_unemployment <- nat_unemployment |>
  filter(!is.na(Month)) |> 
  mutate(Year = floor(Month/100)) |> 
  group_by(Year) |>
  summarize(annual_avg_unemployment = mean(Total, na.rm = T))
annual_avg_unemployment

election_unemployment <- 
  ggplot(data = annual_avg_unemployment,
         mapping = aes(x = Year, y = annual_avg_unemployment)) + geom_smooth(se = FALSE) +
  geom_point() +
   scale_x_continuous(breaks = c(2004, 2008, 2012, 2016, 2020, 2024)) +
  labs(
    title = "Yearly Average Unemployment Rate From Sept 2004-Sept 2024",
    x = "Presidential Election Year",
  y = "Average Unemployment Rate",
caption = "U.S. Bureau of Labor Statistics 2024" +
  theme_minimal(base_size = 14))
election_unemployment

```

# Pivot County Tibble
```{r}

library(ggplot2)
library(tidyr)

county_results <- read_csv("county_resultes - Sheet1.csv")

county_results <- county_results |>
  mutate(`MOV 2020` = case_when(
    `Status 2020` == "Boomerang" ~ `MOV 2020`, 
    `Status 2020` == "Retained" ~ - `MOV 2020`,
    TRUE ~ `MOV 2020`))
county_results
 
county_results_long <- county_results |>
  pivot_longer(
    cols = c("Trump MOV 2016", "Obama MOV 2012", "Obama MOV 2008", "MOV 2020"),
    names_to = "Year",
    values_to = "MOV"
) |>
  mutate(
    Year = as.character(Year), 
    Year = case_when(
      Year == "MOV 2020" ~ 2020,
      Year == "Trump MOV 2016" ~ 2016,
      Year == "Obama MOV 2012" ~ 2012,
      Year == "Obama MOV 2008" ~ 2008
    ),
    "Status 2020" = factor(`Status 2020`, levels = c("Retained", "Boomerang")))


# creating a facet graph for each election
facet_plot <- ggplot(data = county_results_long, 
                     aes(x = County, y = MOV, fill = `Status 2020`)) +
  geom_col() +
  facet_wrap(~ Year) + 
  scale_fill_manual(values = c("Retained" = "red", "Boomerang" = "blue")) + 
  labs(
    title = "Margins of Victory by County and Year",
    x = "County",
    y = "Margin of Victory (%)",
    fill = "Status 2020"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  
    strip.text = element_text(size = 12))
facet_plot

```

## Linear Regression and P value
```{r}
county_results_long$Year <- as.numeric(county_results_long$Year)
annual_avg_unemployment$Year <- as.numeric(annual_avg_unemployment$Year)

combined_data <- left_join(county_results_long, annual_avg_unemployment, by = "Year")
combined_data

impactmodel <- lm(MOV ~ annual_avg_unemployment * `Status 2020`, data = combined_data)
summary(impactmodel)
```



## extras, just showing set seed skills and sample 


```{r}
set.seed(123)

Pivot_Sample <- county_results |>
  sample_n(30)  

county_results <- Pivot_Sample |>
  gather(key = "Election_year", value = "MOV", 
         `Obama MOV 2008`, `Obama MOV 2012`, `Trump MOV 2016`, `MOV 2020`)

Pivot_Sample <- county_results |>
  mutate(Candidate = case_when(
    str_detect(Election_year, "Obama") ~ "Obama",
    str_detect(Election_year, "Trump") ~ "Trump",
    TRUE ~ "Biden"
  ))

  ggplot(Pivot_Sample, aes(x = Election_year, y = County, fill = MOV)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "hotpink", mid = "pink", midpoint = 11, name = "Percent MOV ") +
  labs(title = "30 Randomly Sampled Pivot Counties: MOV Heatmap by Candidate",
       x = "Election Year", y = "County", caption = "Data Source: BALLOTPEDIA") + 
  facet_wrap(~ Candidate) + 
  theme_minimal()


```

  
  
  
  
  
  
  
  
  
  
  