---
title: "PS270 Final Project"
author: "Arwen Lyons "
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(error = TRUE)
```

## Project thoughts (Milestone 1)

I am interested in exploring data related to potentially gerrymandering statistics, potentially picking one county in a swing state that has gone a bunch of different ways and trying to make some sense of that data. I know that some counties voted for Obama then Trump then Biden, so maybe something to do with voting statistics in those counties. Otherwise prison gerrymandering might also be interesting to look into. 


## Final Project Milestone 2

I am curious to look at the relationship between unemployment rates on and around election day and presidential election results in pivot counties across the US. 

I will average the unemployment rates for the month prior to election day to account for early and absentee voting. 

For my dependent variable (democrat or republican victory) I will use the data set from ballotpedia, for my independent variable (unemployment rate), I will use the data set from the US Bureou of Labor Statistics. 

My tentative research question is: Using voting records from the 206 "Pivot Counties" do high rates of unemployment leading up to a presidential election disproportionately harm the republican or democratic party? 

I hypothesize that the economic indicator of high unemployment rates will disproportionately result in greater margins of republican victory, the higher the unemployment rate, the larger the republican margin of victory will be. My hypothesis is based on the polling that has shown Americans trust republicans more than democrats on handling of the economy (regardless or actual economic growth under a republican or democratic president) I think many voters particularity in these pivot counties who vote less reliably on party lines are influenced heavily by their perceptions of economic strength, unemployment is one of the simplest and most telling economic indicators. Both of my variables are quantitative so I won't have to measure them on a sliding scale or by grouping factors, just margins of republican or democratic victory by election year n each of these pivot counties compared to national unemployment rates. 

The observed pattern in this data that would support my hypothesis will be that in years with higher unemployment rates pivot counties across the united sates swing republican by greater margins. If pivot counties vary drastically in margins of republican victory regardless of unemployment rates this would disprove my hypothesis or if democrats perform better or the same with high unemployment rates. 

Resources:
https://www.bls.gov/cps/
https://ballotpedia.org/Election_results,_2020:_Pivot_Counties_in_the_2020_presidential_election
https://www.nbcnews.com/news/amp/rcna117054


## Final Project Milestone 3
This milestone requires that your website loads the data you have selected and produces one interesting and polished data visualization. This could either show the distribution of one variable or the relationship between two variables.

 -------------------- ----------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
`County`              Name of Pivot County 

`state`               State where county resides

`status.2020`         labeled retained when counties kept their voting record,
                      ie: Trump in 2016 and Trump in 2020 or Trump in 2016 and 
                      Biden in 2020, counties that switched votes are labeled
                      boomerang

`MOV.2020`            The winning parties margin of victory in 2020

`Trump.MOV.2016`      Republican Candidate Donald Trump's margin of victory in
                      2016

`Obama.MOV.2012`      Democratic presidential incumbent Obama's margin of 
                      victory in 2012

`Obama_MOV.2008`      Demoratic Canidate Obama's margin of vicotry in 2008


(Need to muthate/ rename) nat_unemployment    National unemployment rate as reported by the Buearo of
                    Labor Statistics in a given election year
 ------------------------------------------------------------------------------
 
# loading csv.files
```{r}
library(tidyverse)
library(broom)
library(lubridate)

county_results <- read_csv("county_resultes - Sheet1.csv")
county_results

nat_unemployment <- read_csv("National Unemployment Statistics sep 2004 - sept 2024 - Sheet1.csv")
nat_unemployment


```



```{r}
nat_unemployment <- as_tibble(nat_unemployment)
nat_unemployment
```

# simplifying data
```{r}

nat_unemployment <- nat_unemployment |>
  mutate(Year = year(mdy(paste0("01 ", Month))))

nat_unemployment |>
select(Month,Total, Year)

```
# graph creation
```{r}
unemployment_rates <- 
  ggplot(data = nat_unemployment,
         mapping = aes(x = Year, y = Total)) + geom_smooth(se = FALSE) +
  geom_point() + 
   scale_x_continuous(breaks = c(2004, 2008, 2012, 2016, 2020, 2024)) +
  labs(
    title = "Average Unemployment Rate, Sept 2004-Sept 2024",
    x = "Presidential Election Year",
  y = "Total Unemployment Rate")

unemployment_rates

```


# Pivot County Tibble
```{r}
county_results

county_results <- as_tibble(county_results) 
county_results
```



## Final Project Milestone 4/ future questions and ideas(I will make an office hour appointment)

then Join data sets?

- in the original published data for the pivot counties BallotPedia uses red and blue in the status column is there a way to add this or what would be the best alternative in r studio to show the same point?

- ideally for my unemployment data set I want to only have the total unemployment rate for 2004, 2008, 2012, 2016, 2020, and 2024 
  - and then do I want to eventually average the months together in all of those years? 

- for my graph, not exactly where I want it I would like to have the months still there and not on a vertical line this is kind of an interesting way of looking at the data but I don't think it is the best for my purposes, I just feel like I don't know enough in r studio to experiment or know hypothetically what the best graph would be.



