---
title: "PS270 Final Project"
author: "Arwen Lyons "
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(error = TRUE)
```

## Project thoughts (Milestone 1)

I am interested in exploring data related to potentially gerrymandering statistics, potentially picking one county in a swing state that has gone a bunch of different ways and trying to make some sense of that data. I know that some counties voted for Obama then Trump then Biden, so maybe something to do with voting statistics in those counties. Otherwise prison gerrymandering might also be interesting to look into. 


## Final Project Milestone 2

I am curious to look at the relationship between unemployment rates on and around election day and presidential election results in pivot counties across the US. 

I will average the unemployment rates for the month prior to election day to account for early and absentee voting. 

For my dependent variable (democrat or republican victory) I will use the data set from ballotpedia, for my independent variable (unemployment rate), I will use the data set from the US Bureou of Labor Statistics. 

My tentative research question is: Using voting records from the 206 "Pivot Counties" do high rates of unemployment leading up to a presidential election disproportionately harm the republican or democratic party? 

I hypothesize that the economic indicator of high unemployment rates will disproportionately result in greater margins of republican victory, the higher the unemployment rate, the larger the republican margin of victory will be. My hypothesis is based on the polling that has shown Americans trust republicans more than democrats on handling of the economy (regardless or actual economic growth under a republican or democratic president) I think many voters particularity in these pivot counties who vote less reliably on party lines are influenced heavily by their perceptions of economic strength, unemployment is one of the simplest and most telling economic indicators. Both of my variables are quantitative so I won't have to measure them on a sliding scale or by grouping factors, just margins of republican or democratic victory by election year n each of these pivot counties compared to national unemployment rates. 

The observed pattern in this data that would support my hypothesis will be that in years with higher unemployment rates pivot counties across the united sates swing republican by greater margins. If pivot counties vary drastically in margins of republican victory regardless of unemployment rates this would disprove my hypothesis or if democrats perform better or the same with high unemployment rates. 

Resources:
https://www.bls.gov/cps/
https://ballotpedia.org/Election_results,_2020:_Pivot_Counties_in_the_2020_presidential_election
https://www.nbcnews.com/news/amp/rcna117054


## Final Project Milestone 3
This milestone requires that your website loads the data you have selected and produces one interesting and polished data visualization. This could either show the distribution of one variable or the relationship between two variables.

 -------------------- ----------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
`County`              Name of Pivot County 

`state`               State where county resides

`status.2020`         labeled retained when counties kept their voting record,
                      ie: Trump in 2016 and Trump in 2020 or Trump in 2016 and 
                      Biden in 2020, counties that switched votes are labeled
                      boomerang

`MOV.2020`            The winning parties margin of victory in 2020

`Trump.MOV.2016`      Republican Candidate Donald Trump's margin of victory in
                      2016

`Obama.MOV.2012`      Democratic presidential incumbent Obama's margin of 
                      victory in 2012

`Obama_MOV.2008`      Democratic Candidate Obama's margin of victory in 2008


`total`               Total National unemployment rate as reported by the Buero of
                      Labor Statistics in a given election year
 ------------------------------------------------------------------------------
 
# loading csv.files
```{r}
library(tidyverse)
library(broom)
library(lubridate)
library(ggplot2)

county_results <- read_csv("county_resultes - Sheet1.csv")
county_results

nat_unemployment <- read_csv("National Unemployment Statistics sep 2004 - sept 2024 - Sheet1.csv")
nat_unemployment


```



```{r}
nat_unemployment <- as_tibble(nat_unemployment)
nat_unemployment
```

# simplifying data
```{r}

nat_unemployment <- nat_unemployment |>
  mutate(Year = year(mdy(paste0("01 ", Month))))
nat_unemployment
```
# Pivot County Tibble
```{r}
county_results

county_results <- as_tibble(county_results) 
county_results
```

# Graph Creation
```{r}
unemployment_rates <- 
  ggplot(data = nat_unemployment,
         mapping = aes(x = Year, y = Total)) + geom_smooth(se = FALSE) +
  geom_point() + 
   scale_x_continuous(breaks = c(2004, 2008, 2012, 2016, 2020, 2024)) +
  labs(
    title = "Average Unemployment Rate, Sept 2004-Sept 2024",
    x = "Presidential Election Year",
  y = "Total Unemployment Rate") +
  theme_minimal()

unemployment_rates
```



## Milestone 4

Your article should contain one visualization and one analysis that attempts to answer your research question. There does not need to be a long discussion, but the results should be presented in either a second visualization or a nicely formatted table.

# Graph creation 
```{r}

unemployment_rates <- 
  ggplot(data = nat_unemployment,
         mapping = aes(x = Year, y = Total)) + geom_smooth(se = FALSE) +
  geom_point() + 
   scale_x_continuous(breaks = c(2004, 2008, 2012, 2016, 2020, 2024)) +
  labs(
    title = "Average Unemployment Rate, Sept 2004-Sept 2024",
    x = "Presidential Election Year",
  y = "Total Unemployment Rate", caption = "Data Source: U.S Bureau of Labor Statistics") +
  theme_minimal()

unemployment_rates


set.seed(123) # is this necessary I know we did this in section and a problem set so it is reproducible but IDK
Pivot_Sample <- county_results |>
  sample_n(30)  


county_results_long <- Pivot_Sample |>
  gather(key = "Election_year", value = "MOV", 
         `Obama MOV 2008`, `Obama MOV 2012`, `Trump MOV 2016`, `MOV 2020`)

Pivot_Sample <- county_results_long |>
  mutate(Candidate = case_when(
    str_detect(Election_year, "Obama") ~ "Obama",
    str_detect(Election_year, "Trump") ~ "Trump",
    TRUE ~ "Biden"
  ))

  ggplot(Pivot_Sample, aes(x = Election_year, y = County, fill = MOV)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "hotpink", mid = "pink", midpoint = 11, name = "Percent MOV ") +
  labs(title = "30 Randomly Sampled Pivot Counties: MOV Heatmap by Candidate",
       x = "Election Year", y = "County", caption = "Data Source: BALLOTPEDIA") + 
  facet_wrap(~ Candidate) + 
  theme_minimal()

```

## Questions for office hours 
- I don't know where the code for labeling the x axis line is cominf from so i was not able to change it, somehow I need to make it so the collums are only for each year not showing the MOV for all three candiates for the same year every collum (obviosly no data for obama MOV in 2020, etc)
  - I don't know if I want all of the extra data for unemployment rates, I could not figure out how to get rid of it and simplify it to just month, year, and total rate 
  
  - along those lines how would I go about separating month and year so they are in a separate column, and do I even want to dot that, or deally for my unemployment data set I want to only have the total unemployment rate for 2004, 2008, 2012, 2016, 2020, and 2024, and then do I want to eventually average the months together in all of those years? 
  
  - - in the original published data for the pivot counties BallotPedia uses red and blue in the status column is there a way to add this or what would be the best alternative in r studio to show the same point?
  
  - I have given up on my attempts to join the data sets because I could not figure out how to it because they do not share any column names and have a different number of columns 
  
